---
title: "Interactive scatterplots with scatterD3"
author: "Julien Barnier"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 5
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `scatterD3` package provides an HTML widget based on the `htmlwidgets` package and allows to produce interactive scatterplots by using the `d3.js` javascript visualization library.

## Basic scatterplot

Starting with the sample `mtcars` dataset, we can produce a basic scatterplot with the following command :

```{r basic}
library(scatterD3)
scatterD3(x = mtcars$wt, y = mtcars$mpg)
```

This will display a simple visualization with the given variables as `x` and `y` axis. There are several interactive feature directly available :

- you can zoom in and out with the mouse wheel while the mouse cursor is on the plot
- while zommed in, you can pan the plot by dragging with your mouse
- by hovering over a point, you can display a small tooltip window giving the `x` and `y` values

You can customize the points size with the `point_size` parameter, their opacity with `point_opacity`, and you can force the plot to have a 1:1 fixed aspect ratio with `fixed=TRUE`.

```{r basic_cust}
scatterD3(x = mtcars$wt, y = mtcars$mpg, point_size = 15, point_opacity = 0.5, fixed = TRUE)
```

## Point labels

You can add text labels to the points by passing a character vector to the `lab` parameter. Labels size are controlled by the `labels_size` parameter.

```{r labels}
scatterD3(x = mtcars$wt, y = mtcars$mpg, lab = rownames(mtcars), labels_size = 9)
```

Note that text labels are fully movable : click and drag a label with your mouse to place it where you want. Custom positions are preserved while zooming/panning.


## Mapping colors and symbols to variables

By passing vectors to the `col_var` and/or `symbol_var` arguments, you can map points colors and symbols to other variables.

```{r mapping}
scatterD3(x = mtcars$wt, y = mtcars$mpg, col_var = mtcars$cyl, symbol_var = mtcars$gear)
```

Note that when hovering over a legend item with your mouse, the corresponding points are highlighted. Also note that the mapped variables values are automatically added to the default tooltips.


## Custom axis and legend labels

You can customize the axis and legend labels with the `xlab`, `ylab`, `col_lab` and `symbol_lab` :

```{r cust_labels}
scatterD3(x = mtcars$wt, y = mtcars$mpg, col_var = mtcars$cyl, symbol_var = mtcars$gear,
          xlab = "Weight", ylab = "Mpg", col_lab = "Cylinders", symbol_lab = "Gears")
```

Note that default tooltips are updated accordingly.


## Custom tooltips

If the default tooltip doesn't suit your needs, you can customize them by providing a character vector to the `tooltip_text`. This can contain HTML tags for formatting.

```{r cust_tooltips}
tooltips <- paste("This is an incredible <strong>", rownames(mtcars),"</strong><br />with ", 
                  mtcars$cyl, "cylinders !")
scatterD3(x = mtcars$wt, y = mtcars$mpg, tooltip_text = tooltips)
```

You can also disable tooltips entirely with `tooltips = FALSE`.

## Shiny integration : additional controls and SVG export

Like every R HTML widget, shiny integration is straightforward. Furthermore, we provide some additional
handlers to map form controls to SVG export and interactive effects (text size, points opacity and zoom reset).
You just have to give the following `id` to your form controls :

- `#scatterD3-size` : text size in points (numerical value)
- `#scatterD3-opacity` : point ant text opacity (numerical value, 0 to 1)
- `#scatterD3-resetzoom` : reset zoom to default on click
- `#scatterD3-download` : link to download the currently displayed figure as an SVG file

Here is a minimal working example :

```R
library(shiny)
runApp(shinyApp(
    ui=fluidPage(
      sidebarLayout(
        sidebarPanel(
          numericInput("scatterD3-size", "Labels size :", min = 2, max = 30, value = 10),
          numericInput("scatterD3-opacity", "Opacity :", min = 0, max = 1, value = 1, step=0.05),
          actionButton("scatterD3-resetzoom", "Reset Zoom"),
          tags$a(id="scatterD3-download", href="#", class="btn btn-default", "Download SVG")
        ),
        mainPanel(scatterD3Output("scatterPlot"))
      )
    ),
    server = function(input, output) {
      output$scatterPlot <- renderScatterD3({
        scatterD3(x=mtcars$wt,
                  y=mtcars$mpg,
                  lab=rownames(mtcars),
                  col_var=mtcars$cyl)
      })
    }
))
```

You can see the result of this [minimal scatterD3 shiny app](https://juba.shinyapps.io/scatterD3_shiny_app) hosted on shinyapps.io.

