---
title: "Interactive scatterplots with scatterD3"
author: "Julien Barnier"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 5
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `scatterD3` package provides an HTML widget based on the `htmlwidgets` package and allows to produce interactive scatterplots by using the `d3.js` javascript visualization library.

## Basic scatterplot

Starting with the sample `mtcars` dataset, we can produce a basic scatterplot with the following command :

```{r basic}
library(scatterD3)
scatterD3(x = mtcars$wt, y = mtcars$mpg)
```

This will display a simple visualization with the given variables as `x` and `y` axis. There are several interactive features directly available :

- you can zoom in and out with the mouse wheel while the mouse cursor is on the plot
- you can pan the plot by dragging with your mouse
- by hovering over a point, you can display a small tooltip window giving the `x` and `y` values

You can customize the points size with the `point_size` parameter, their opacity with `point_opacity`, and you can force the plot to have a 1:1 fixed aspect ratio with `fixed = TRUE`.

```{r basic_cust}
scatterD3(x = mtcars$wt, y = mtcars$mpg, point_size = 15, point_opacity = 0.5, fixed = TRUE)
```

## Point labels

You can add text labels to the points by passing a character vector to the `lab` parameter. Labels size are controlled by the `labels_size` parameter.

```{r labels}
scatterD3(x = mtcars$wt, y = mtcars$mpg, lab = rownames(mtcars), labels_size = 9)
```

Note that text labels are fully movable : click and drag a label with your mouse to place it where you want. Custom positions are preserved while zooming/panning.


## Mapping colors, symbols and size to variables

By passing vectors to the `col_var` and/or `symbol_var` arguments, you can map points colors and symbols to other variables.

```{r mapping}
scatterD3(x = mtcars$wt, y = mtcars$mpg, col_var = mtcars$cyl, symbol_var = mtcars$gear)
```

Note that when hovering over a legend item with your mouse, the corresponding points are highlighted. Also note that the mapped variables values are automatically added to the default tooltips.

You can also map symbol sizes with a variable with the `size_var` argument. `size_range` allows to customize the sizes range :

```{r map_size}
scatterD3(x = mtcars$wt, y = mtcars$mpg, col_var = mtcars$cyl, size_var = mtcars$hp, 
          size_range = c(10,1000), point_opacity = 0.7)
```

## Axis limits

You can manually specify the `x` or `y` axis limits with the `xlim` and `ylim` arguments :

```{r axis_limits}
scatterD3(x = mtcars$wt, y = mtcars$mpg, xlim=c(0,10), ylim=c(10,35))
```



## Custom axis and legend labels

You can customize the axis and legend labels with `xlab`, `ylab`, `col_lab`, `symbol_lab` and `size_lab` :

```{r cust_labels}
scatterD3(x = mtcars$wt, y = mtcars$mpg, col_var = mtcars$cyl, symbol_var = mtcars$gear,
          xlab = "Weight", ylab = "Mpg", col_lab = "Cylinders", symbol_lab = "Gears")
```

Note that default tooltips are updated accordingly.


## Custom tooltips

If the default tooltips don't suit your needs, you can customize them by providing a character vector to the `tooltip_text` argument. This can contain HTML tags for formatting.

```{r cust_tooltips}
tooltips <- paste("This is an incredible <strong>", rownames(mtcars),"</strong><br />with ", 
                  mtcars$cyl, "cylinders !")
scatterD3(x = mtcars$wt, y = mtcars$mpg, tooltip_text = tooltips)
```

You can also disable tooltips entirely with `tooltips = FALSE`.

## Other options

Finally, and for more specific use cases, you can represent some points as an arrow starting from the origin by using the `type_var` argument, and you can add a unit circle with `unit_circle = TRUE`.

```{r cust_arrows}
scatterD3(x = c(1, 0.9, 0.7, 0.2, -0.4, -0.5), xlab = "x",
          y = c(1, 0.1, -0.5, 0.5, -0.6, 0.7), ylab = "y",
          lab = LETTERS[1:6], type_var = c("point", rep("arrow", 5)),
          unit_circle = TRUE, fixed = TRUE, xlim = c(-1.2, 1.2))
```


## Shiny integration

### Transitions

Like every R HTML widget, shiny integration is straightforward. But as a D3 widget, `scatterD3` is *updatable* : changes in settings or data can be displayed via smooth transitions instead of a complete chart redraw, which can provide interesting visual clues.

For a small demonstration of these transitions, you can take a look at the
[sample scatterD3 shiny app](http://data.nozav.org/app/scatterD3/).

Enabling transitions in your shiny app is quite simple, you just have to add the `transitions = TRUE` argument to your `scatterD3` calls in your shiny server code. There's only one warning : if your shiny application may filter on your dataset rows via a form control, then you must provide a `key_var` variable that uniquely and persistently identify your rows.


### Additional controls : Reset zoom and SVG export

Furthermore, `scatterD3` provides some additional handlers for two interactive features : SVG export and zoom resetting.

By default, you just have to give the following `id` to the corresponding form controls :

- `#scatterD3-reset-zoom` : reset zoom to default on click
- `#scatterD3-svg-export` : link to download the currently displayed figure as an SVG file

If you are not happy with these ids, you can specify their names yourself with the arguments `dom_id_svg_export` and `dom_id_reset_zoom`.

### Sample app and source code

The
[sample scatterD3 shiny app](http://data.nozav.org/app/scatterD3/) allows you to see the different features described here. You can [check its source code on GitHub](https://github.com/juba/scatterD3_shiny_app) for a better understanding of the different arguments.
